---
title: "Clustering"
output: html_notebook
---

```{r project-paths, include=FALSE}
source(if (file.exists("R/project_paths.R")) "R/project_paths.R" else "../R/project_paths.R")
set_project_root()
```

```{r}
clocks <- c("McEwen", "Wu", "Hannum", "Horvath", "Levine", "Belsky", "Teschendorff", "Shireby", "Lu")
selected_clock_cpgs <- readRDS("relevant_rds/clock_cpgs.rds")
if (!length(selected_clock_cpgs)) {
  stop("The filtered clock CpG list is empty.")
}

# The clock-membership annotation is constructed inside mainPipeline after the
# matrix has been filtered. This guarantees that its rows stay aligned with the
# heatmap and avoids relying on a separately saved row-order file.
```

```{r}
# params: num_clusters, SAVE, CLOCK, gender, ret, labs

# param ret: either "df" to get the matrix with clusters annotated for use in further pipeline, or "plot" to get the plot, for use in combining in Fig 2 and 5
# param labs: only relevant if ret == "plot", just specifying what labels you want

# Each call returns both the cluster assignments and its heatmap/scatter panel,
# so the expensive clustering step runs only once per cohort.
non_run <- mainPipeline(4, FALSE, FALSE, "all", "both", c("C", "D"))
clock_run <- mainPipeline(4, FALSE, TRUE, "all", "both", c("A", "B"))
male_run <- mainPipeline(4, FALSE, FALSE, "male", "both", c("A", "B"))
female_run <- mainPipeline(4, FALSE, FALSE, "female", "both", c("C", "D"))

non_all <- non_run$data
clock_all <- clock_run$data
male_all <- male_run$data
female_all <- female_run$data

fig2a <- clock_run$plot
fig2b <- non_run$plot
fig5a <- male_run$plot
fig5b <- female_run$plot

cluster_output_dir <- getOption(
  "agepattern.cluster_output_dir",
  "relevant_rds"
)
dir.create(cluster_output_dir, recursive = TRUE, showWarnings = FALSE)
saveRDS(non_all, file.path(cluster_output_dir, "all_cluster_assignments.rds"))
saveRDS(clock_all, file.path(cluster_output_dir, "clock_cluster_assignments.rds"))
saveRDS(male_all, file.path(cluster_output_dir, "male_cluster_assignments.rds"))
saveRDS(female_all, file.path(cluster_output_dir, "female_cluster_assignments.rds"))


```

Determine enrichment of CpG sites in clusters
- enrichment of clock CpGs in clock clusters
- enrichment of aaCpGs in each window in aaCpG clusters (Figure 2F)

*Necessary to run mainPipeline to have the matrices with cluster annotations ready, but can be done quickly by specifying save=FALSE and plot_smooth=FALSE
*Necessary to run filter code to have list1 ready

```{r}

# Run

ht <- clusterEnrichment(clock_all, 4, clock_cpgs_list)
p3 <- grid.grabExpr(draw(ht, column_title = " ", column_title_gp = gpar(fontsize=14, fontface="bold", fontfamily="Helvetica")))

ht2 <- clusterEnrichment(non_all, 4, all_list)
p4 <- grid.grabExpr(draw(ht2, column_title = " ", column_title_gp = gpar(fontsize=14, fontface="bold", fontfamily="Helvetica")))

fig2c <- plot_grid(p3, p4, labels=c("E", "F"), rel_widths=c(0.8, 1.2))
```

```{r}
# combined for fig2 -> fig2a, fig2b, p3, and p4

combined <- plot_grid(fig2a, fig2b, fig2c, nrow=3, rel_heights = c(1, 1, 0.8))
output_dir <- getOption("agepattern.output_dir", "new_figs")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
ggsave(file.path(output_dir, "figure2.pdf"), plot = combined, device=cairo_pdf,
        width = 15, height = 16, dpi=300)

# combined for fig5 <- fig5a, fig5b

combined <- plot_grid(fig5a, fig5b, nrow=2)
ggsave(file.path(output_dir, "figure5.pdf"), plot = combined, device = cairo_pdf, width = 15, height = 11.42, dpi=300)


```



UNUSED / RETAINED FOR REFERENCE BELOW

The remaining chunks are exploratory genomic-distribution work and are not
required to reproduce the primary clustering figures.

```{r}
library(circlize)
library(gridBase)

ann <- read.table("data/ref/HM450.hg38.manifest.tsv", sep="\t", header=TRUE, quote="", comment.char="", fileEncoding = "UTF-8")
ann <- ann[!is.na(ann$CpG_chrm) & !is.na(ann$CpG_beg) & !is.na(ann$CpG_end), ]
rownames(ann) <- ann$probeID
#earlyagegroup <- four_cluster_select[four_cluster_select$Cluster==1 | four_cluster_select$Cluster==3, ]
#earlycpgs <- rownames(earlyagegroup)

#oldagegroup <- four_cluster_select[four_cluster_select$Cluster==2 | four_cluster_select$Cluster==4, ]
#oldcpgs <- rownames(oldagegroup)
all_aaCpGs <- readRDS("relevant_rds/all_select_cpg.rds")
male_aaCpGs <- readRDS("relevant_rds/male_select_cpg.rds")
female_aaCpGs <- readRDS("relevant_rds/female_select_cpg.rds")

K450 <- ann[, 1:3]
colnames(K450) <- c("chr", "start", "end")
selected <- K450[rownames(K450) %in% all_aaCpGs, ]
maleSelected <- K450[rownames(K450) %in% male_aaCpGs, ]
femaleSelected <- K450[rownames(K450) %in% female_aaCpGs, ]
#youngSelected <- K450[rownames(K450) %in% earlycpgs, ]
#oldSelected <- K450[rownames(K450) %in% oldcpgs, ]
#all_sites <- readRDS("7_23_all_sites.rds")
#K300 <- ann[rownames(ann) %in% all_sites, 1:3]

#set.seed(123)
#M1 <- M28[sample(nrow(M28), size=1000000, replace=FALSE), ]



lgd_points = Legend(at = c("450K CpGs"), type = "points", 
    legend_gp = gpar(col = "blue"), title_position = "topleft", 
    title = " ")
lgd_points2 = Legend(at = c("Selected aaCpGs"), type = "points", 
    legend_gp = gpar(col = "red"), title_position = "topleft", 
    title = " ")
lgd_points3 = Legend(at = c("Male aaCpGs"), type = "points", 
    legend_gp = gpar(col = "green"), title_position = "topleft", 
    title = " ")
lgd_points4 = Legend(at = c("Female aaCpGs"), type = "points", 
     legend_gp = gpar(col = "purple"), title_position = "topleft", 
     title = " ")
lgd_list_vertical = packLegend(lgd_points, lgd_points2, lgd_points3, lgd_points4)

#save(M28, K300, selected, file="10_15.RData")
#load("10_15.RData")

png("new_figs/temp_circos.png", width = 8, height=6, units="in", res=300)
plot.new()
circle_size = unit(1, "snpc")
pushViewport(viewport(x = 0, y = 0.5, width = circle_size, height = circle_size,
    just = c("left", "center")))
par(omi = gridOMI(), new = TRUE)
#vps <- baseViewports()
#par(new = TRUE, fig = gridFIG())
circos.initializeWithIdeogram(chromosome.index = paste0("chr", 1:22), species="hg38")
        #circos.genomicDensity(M28, col = c("#FF000080"), track.height = 0.1)
circos.genomicDensity(K450, col=c("blue"), track.height=0.1)
circos.genomicDensity(selected, col=c("red"), track.height=0.1)
circos.genomicDensity(maleSelected, col=c("green"), track.height=0.1)
circos.genomicDensity(femaleSelected, col=c("purple"), track.height=0.1)

upViewport()
draw(lgd_list_vertical, x = circle_size, just = "left")
dev.off()
```
